---
title: "Bey2ollak Traffic Analaysis"
author: "Mohamed Farghal"
date: "7 April 2016"
output: html_document
---

The Traffic dataset is collected from bey2ollak.com, This EDA summarizes some of important data features and show some of the traffic activities and patterns in Cairo.  


```{r, warning=FALSE}
library(grid)
library(ggplot2)
library(dplyr)
library(mixtools)
```

```{r, echo=FALSE}
Sys.setlocale('LC_ALL', 'en_US.UTF-8')
```


```{r}
data <- read.csv("../traffic-data.csv", encoding="UTF-8")
data %>% names()
```

```{r}
data %>% nrow()
```


### Data Summary

```{r}
data %>% str()
```

```{r}
data %>% summary()
```

Now we need to define the report time.

Let's have a look on rd.rp.hr.

```{r}
data$rd.rp.hr %>% boxplot()
```

rd.rp.hr has some data above 24. Thus, we can confirm that it is the hours elapsed.


The rows represent a traffic report made on the website. The columns meaning can be extracted by looking into the website and how they use the data to render the fields. The meaning of the important columns are:

Column Pattern | Meaning
-------------- | --------------
crawl_date     | time data was crawled
ad.*           | information related to advertisment
rd.*           | information related to a road
hr             | the number of hours elapsed from the report was made to crawling date
mn             | hr + mn gives the exact time elapased from the report date to crawling date
rd.rp          | data related to a report
rd.rp.fullnm   | reporter fullname
rp.rp.nm       | reporter username
stid           | rd.rp.stid is the status id of the report; rd.stid is the road status id
rd.nm          | road name (<name>;<source> To <Destination>)
rd.rp.cm       | user comment
rd.ri          | road id


Now let's calcualte the actual report time.

```{r}
data$crawl_date         <- as.POSIXct(strptime(data$crawl_date, format="%A %b %d %H:%M:%S UTC %Y"), tz='UTC')
data$crawl_date         <- as.POSIXlt(data$crawl_date, tz="Africa/Cairo")
data$report_date        <- data$crawl_date
data$report_date$hour   <- (data$crawl_date$hour - data$rd.rp.hr) %% 24
data$report_date$minute <- data$crawl_date$minute - data$rd.rp.mn
```

Now we prepare Time units for usage simplicity.

```{r}
data$hour <- data$report_date$hour
data$min  <- data$report_date$min
data$wday <- data$report_date$wday
data$yday <- data$report_date$yday
```

The stid columns hold report status. Let's confirm the meaning of each id by viewing a sample of the comments for each status id.

stid = 1
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 1, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 2
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 2, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 3
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 3, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 4
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 4, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 5
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 5, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 6
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 6, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 7
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 7, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 8
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 8, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 9
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 9, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

stid = 10
```{r}
data[!is.na(data$rd.rp.stid) & data$rd.rp.stid == 10, c('rd.rp.stid', 'rd.rp.cm')] %>% head(10)
```

According to the data samples and the label on the website. the stid categories are:
```{r}
stid <- c("7alawa", "lazez", "mashy" , "za7ma", "Noamal", "Question", "Khatar", "7adsa", "3otl", "ma3loma")
```


### Data Cleaning and Feature Engineering

First we need to remove duplicate rows

```{r}
data.unq <- data[!duplicated(data[, c('rd.rp.cmid')]), ]
```


Weekends. Friday and Saturday are the weekends in Egypt. Let's use this to define a weekend row.

```{r}
data.weekend <- data.unq[data.unq$wday >= 5, ]
data.workday <- data.unq[data.unq$wday < 5, ]
```


We can add labels for for the city (alex or cairo) which is retrieved from an XML file from the website. The XML file contains road name and road id which I parsed with a script that saves the ids in the following csv files.

```{r}
cairo.roads <- read.csv("cairo-roads-ids.csv")
alex.roads <- read.csv("alex-roads-ids.csv")
cairo.roads$city = "cairo"
alex.roads$city = "alex"
data.unq <- merge(data.unq, rbind(cairo.roads, alex.roads), by="rd.ri")
```


Road names are written in a certian format as shown in the above columns definition. 

```{r}
data.unq$rd.nm %>% sample(10)
```

We can decouple the information from the road name to constrcut the source and the destination of the road.

```{r}
data.unq$rd.nm <- as.character(data.unq$rd.nm)
data.unq$road.name <- sapply(data.unq$rd.nm, function(x){unlist(strsplit(x, ";"))[1]})
data.unq$road.direction <- sapply(data.unq$rd.nm, function(x){unlist(strsplit(x, ";"))[2]})
data.unq$road.direction.from <- sapply(data.unq$road.direction, function(x){unlist(strsplit(x, "To"))[1]})
data.unq$road.direction.to <- sapply(data.unq$road.direction, function(x){unlist(strsplit(x, "To"))[2]})
data.unq$road.direction.from.full <- sapply(data.unq$rd.nm, function(x){unlist(strsplit(x, "To"))[1]})
```


```{r}
data.unq[, c("road.direction.from", "road.direction.to", "road.direction.from.full")] %>% unique() %>% head(10)
```

Let's have a look on the crawling distribtion per hour.

```{r}
qplot(data$crawl_date$hour, geom="histogram", binwidth=1, xlab = "Crawling Hour", ylab = "Count")
```

The distribution of the crawling hour cannot tell accurate information about activities.

Let's make sure that the crawling was made continously without breaks.

```{r}
ggplot(data, aes(yday)) + stat_ecdf(geom = "step") + xlab("Year Day") + ylab("CMF")
```

Luckily, There's no obvious breaks.


Let's focus on the important columns with respect to the scope of traffic analaysis.

```{r}
names.imp <- names(data.unq)
names.imp <- names.imp[!grepl("^ad", names.imp)]
data.unq <- data.unq[, names.imp]
names.imp
```


```{r}
data.unq %>% nrow()
```

Percentage of duplicate data

```{r}
1 - (nrow(data.unq) / nrow(data))
```


### Activity Analaysis

Let's have a look on the activities (reports) patterns and histograms.


Total number of activities per hour

```{r}
qplot(data.unq$hour, geom="histogram", binwidth=1, xlim=c(0,24), xlab='Day hour', ylab='Reports Count')
```

Activities per day
```{r}
qplot(data.unq$yday, geom="histogram", binwidth=1, xlab='Day', ylab='Reports Count')
```


Activities per Weekday

```{r}
qplot(data.unq$wday, geom="histogram", binwidth=1, xlim=c(0,7), xlab='Week day', ylab='Reports Count')
```

Activities per Weekdays

```{r}
qplot(data.workday$hour, geom="histogram", binwidth=1, xlim=c(0,24), xlab='Workday Day hour', ylab='Number of reports')
```

Count of each type of activity

```{r}
p <- ggplot(data.unq , aes(as.factor(rd.rp.stid)))
p + geom_bar(fill=c("#178B37", "#7CBD33", "#FEC00D", "#F37F1B", "#E7252D", "black", "gray", "gray", "gray", "gray", "gray")) + scale_x_discrete(labels=c(stid, "NA"))
```


Count of each type of activity in Cairo only

```{r}
p <- ggplot(data.unq[data.unq$city=="cairo",] , aes(as.factor(rd.rp.stid)))
p + geom_bar(fill=c("#178B37", "#7CBD33", "#FEC00D", "#F37F1B", "#E7252D", "black", "gray", "gray", "gray", "gray", "gray")) + scale_x_discrete(labels=c(stid, "NA"))
```

Count of each type of activity in Cairo only

```{r}
p <- ggplot(data.unq[data.unq$city=="alex",] , aes(as.factor(rd.rp.stid)))
p + geom_bar(fill=c("#178B37", "#7CBD33", "#FEC00D", "#F37F1B", "#E7252D", "black", "gray", "gray", "gray", "gray", "gray")) + scale_x_discrete(labels=c(stid, "NA"))
```

Activities in alex is not huge and is biased, which does not follow the usual patterns.


```{r, result="hide", echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```




## Traffic Analaysis

Road status Id indicate the traffic status of the road. we can use it to have some insights on traffic pattern. 


Let's see what is the traffic pattern per day, by plotting the stid per hour and fit a smoothing line to indicate where the is density per hour.


```{r}
traffic <- data.unq[data.unq$rd.rp.stid < 6,]
traffic <- traffic[, c("rd.rp.stid", "hour", "city")]
traffic <- na.omit(traffic)
scatter.smooth(traffic$hour, traffic$rd.rp.stid, col="#CCCCCC", span = 0.23)
```

Now Let's compare the traffic pattern for Cairo and Alex

```{r}
traffic.cairo <- traffic[traffic$city == "cairo", ]
traffic.alex <- traffic[traffic$city == "alex", ]
p1 <- scatter.smooth(traffic.cairo$hour, traffic.cairo$rd.rp.stid, col="#CCCCCC", span = 0.3)

```

Pattern in Alex is not significant.

```{r}
p2 <- scatter.smooth(traffic.alex$hour, traffic.alex$rd.rp.stid, col="#CCCCCC", span = 0.3)
```


Traffic in cairo is similar to a bimodal distribution where the means can indicate the rush hours.

Let's first define a heavy vs non-heavy traffic

```{r}
data.unq$report.is.za7ma <- data.unq$rd.rp.stid > 3 & data.unq$rd.rp.stid < 6
data.unq$report.is.mshza7ma <- data.unq$rd.rp.stid <= 3
data.unq$road.is.za7ma <- data.unq$rd.stid > 3 & data.unq$rd.stid < 6 & data.unq$rd.hr < 1 
data.unq$road.is.mshza7ma <- data.unq$rd.stid <= 3 & data.unq$rd.hr < 1
```

Then we group by hour to see the number of heavy traffic reports vs non heavy.

```{r}
road.dist <- data.unq[, c("hour", "road.is.za7ma", "road.is.mshza7ma", "city", "report.is.za7ma", "report.is.mshza7ma")] %>%
filter(!is.na(road.is.za7ma)) %>% filter(!is.na(report.is.za7ma)) %>%
group_by(hour) %>%
summarize(za7ma=sum(road.is.za7ma), mshza7ma=sum(road.is.mshza7ma), za7ma.per=za7ma/mshza7ma,
         report.za7ma=sum(report.is.za7ma), report.mshza7ma=sum(report.is.mshza7ma),
         report.za7ma.per=report.za7ma/(report.za7ma + report.mshza7ma))
```


```{r}
p <- ggplot(data=road.dist) + ylab("Reports Count")
p <- p + geom_bar(aes(x=0:23, y=report.mshza7ma), fill="#66CC99", stat="identity" )
p <- p + geom_bar(aes(x=0:23, y=report.za7ma), fill="#FF9999", stat="identity")
p
```

If a road has traffic jam, many reports could be reporting this. Thus, let's have a look on the road status. But we need to make sure that the last report was made less than one hour ago, which was done while setting the road status above.

```{r}
p <- ggplot(data=road.dist) + ylab("Reports Count")
p <- p + geom_bar(aes(x=0:23, y=mshza7ma), fill="#66CC99", stat="identity")
p <- p + geom_bar(aes(x=0:23, y=za7ma), fill="#FF9999", stat="identity")
p <- p + geom_line(aes(x=0:23, y=za7ma, fill="black"), alpha=0.3)
p
```

Perfect. The road status follows the same bimodal pattern. But is this the case in all days?

Let's draw the same plot for each weekday

```{r}
road.day.dist <- data.unq[, c("hour", "wday", "road.is.za7ma", "road.is.mshza7ma", "city", "report.is.za7ma", "report.is.mshza7ma")] %>%
filter(!is.na(road.is.za7ma)) %>% filter(!is.na(report.is.za7ma)) %>%
group_by(hour, wday) %>%
summarize(za7ma=sum(road.is.za7ma), mshza7ma=sum(road.is.mshza7ma), za7ma.per=za7ma/mshza7ma,
         report.za7ma=sum(report.is.za7ma), report.mshza7ma=sum(report.is.mshza7ma),
         report.za7ma.per=report.za7ma/(report.za7ma + report.mshza7ma))
```


```{r}
p <- ggplot(data=road.day.dist) 
p <- p + facet_grid(~wday)
p <- p + geom_bar(aes(x=hour, y=mshza7ma), fill="#66CC99", stat="identity")
p <- p + geom_bar(aes(x=hour, y=za7ma), fill="#FF9999", stat="identity")
p 
```

We can see that all days follow the same distribution excepet for weekends.

We can confirm the distribution by checking the heavy traffic percentage out of all traffic reports.

```{r}
p <- ggplot(data=road.dist)
p <- p + geom_bar(aes(x=0:23, y=report.za7ma.per), fill="orange", stat="identity")
p <- p + geom_line(aes(x=0:23, y=report.za7ma.per, fill="black"), alpha="0.5")
p
```

Now we can assume that the traffic follows a biomodal distriubiton.

Let's focus on cairo and try to pick the two means (peaks) of the traffic by fitting a bimodal distribtuion.

```{r}
library(mixtools)
#Fit two Normals
dist = normalmixEM(data.unq$hour[!is.na(data.unq$rd.rp.stid) & data.unq$city=="cairo" & data.unq$rd.rp.stid > 3 & data.unq$rd.rp.stid < 6], lambda = 0.5)
plot(dist, density=TRUE, loglik=FALSE)
```

Thus the two means are

```{r}
dist$mu
```

with standard deviation of

```{r}
dist$sigma
```

It's also useful to check the mean of the 

Now let's prove that average heavy traffic hour is different on weekdays from weekends

first, the mean and confidence level of the heavy traffic on weekdays

```{r}
road.day.dist$working.day <- road.day.dist$wday < 5
working.dist <- road.day.dist[road.day.dist$working.day == T,]
m <- weighted.mean(working.dist$hour, working.dist$za7ma)
N <- length(working.dist)
variance <- sum(working.dist$za7ma * ((working.dist$hour - m) ^ 2)) / ((sum(working.dist$za7ma) * (N- 1)) / N)  
deviation <- sqrt(variance)
```

Mean and standard deviation

```{r}
paste(m, deviation)
```

Thus, the 95% confidence internval

```{r}
error <- qnorm(0.95)*deviation/sqrt(N)
left <- m - error
right <- m + error
```

The mean of za7ma roads daily hour in workday is with 95% confidence within:

```{r}
paste(left, right)
```

Let's do the same for Weekends.

```{r}
road.day.dist$weekend <- road.day.dist$wday >= 5
weekend.dist <- road.day.dist[road.day.dist$weekend == T,]
m <- weighted.mean(weekend.dist$hour, weekend.dist$za7ma)
N <- length(weekend.dist)
variance <- sum(weekend.dist$za7ma * ((weekend.dist$hour - m) ^ 2)) / ((sum(weekend.dist$za7ma) * (N- 1)) / N)  
deviation <- sqrt(variance)
```

```{r}
paste(m, deviation)
```

```{r}
error <- qnorm(0.95)*deviation/sqrt(N)
left <- m - error
right <- m + error
```

The mean of za7ma roads daily hour in weekends is with 95% confidence within:

```{r}
paste(left, right)
```

This shows confirms that roads in weekends tend to be crowded in later hours than in workdays.

```{r}
qqplot(working.dist$za7ma, weekend.dist$za7ma); qqline(working.dist$za7ma, weekend.dist$za7ma)
```

We can see from the plot that the weekend distribution slightly exceeds the workdays in later hours however, in earlier hour the weekdays exceeds.

### Accidents

Let's have a look on how the accidents in Cairo look like.

```{r}
accidents.hour <- data.unq$hour[data.unq$city == "cairo" & data.unq$rd.rp.stid == 8]
accidents.hour <- na.omit(accidents.hour)
```

Accidents

```{r}
qplot(accidents.hour, geom="histogram", binwidth=1, xlab = "Hour", ylab = "Count")
```

The distribution is similar to the bimodal distribution.

```{r}
dist = normalmixEM(accidents.hour, lambda = 0.5)
plot(dist, density=TRUE, loglik=FALSE)
```

Means of Accidents Hour

```{r}
dist$mu
```


Standard Deviation

```{r}
dist$sigma
```


### Road names

Let's see if we can use the road names to build a heatmap.

```{r}
library("igraph")
q <- data.unq %>% select(road.direction.from, road.direction.to, road.name, city) %>%
  filter(city == "cairo") %>% group_by(road.direction.to, road.direction.from) %>% summarize(count=length(road.name))

q <- q[complete.cases(q), ]

mat <- matrix(0, length(unique(q$road.direction.from)), length(unique(q$road.direction.to))) 
rownames(mat) <- unique(q$road.direction.from)
colnames(mat) <- unique(q$road.direction.to)
f <- as.data.frame(mat)
for (i in 1:length(f)) {
  f[q$road.direction.from[i], q$road.direction.to[i]] <- q$count[i]
}
mat <- as.matrix(f)
palf <- colorRampPalette(c("gold", "blue"))
heatmap(mat, Rowv = NA, Colv = NA, col = palf(100), scale="none", margins=c(10,10) )
```

Road names are not well structured to build a graph from a map.


## Comments Analysis

Let's see if there are significant comments people write when there is accidents.

```{r}
#ignore gps comments
data.unq$rd.rp.cm <- as.character(data.unq$rd.rp.cm)
cm.words <- as.factor(unlist(lapply(data.unq$rd.rp.cm[data.unq$rd.rp.nm != "bey2ollakgps"], function(s){strsplit(s, " ")})))
cm.words <- as.factor(unlist(lapply(data.unq$rd.rp.cm[data.unq$rd.rp.stid == 8], function(s){strsplit(s, " ")})))
cm.words.smpl <- cm.words[sample(500000)]

```

Show a sample of the most common words

```{r}
words.count <- tapply(as.factor(cm.words.smpl), as.factor(cm.words.smpl), FUN=length)
words.count <- sort(words.count, decreasing=T)
words.count %>% head(50)
```

These words might give a hint for the reason of the accident if clustered. for example works like  ???????????? , ?????? , ???????????? might indicate hints about the accidents.

These are a sample of the most common words in heavy traffic reports.

```{r}
m.words <- as.factor(unlist(lapply(data.unq$rd.rp.cm[data.unq$rd.rp.stid == 5], function(s){strsplit(s, " ")})))
cm.words.smpl <- cm.words[sample(5000000)]
words.count <- tapply(as.factor(cm.words.smpl), as.factor(cm.words.smpl), FUN=length)
words.count <- sort(words.count, decreasing=T)
words.count %>% head(50)
```

This shows that people may report an accident with an stid=5



